---
title: "Take-home Exercise 3: Be Weatherwise or Otherwise"
date: "February 14, 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
editor: visual
---

## 1. Overview

The objective of this exercise is to create an analytics-driven data visualization to validate the claims that daily mean temperatures are projected to increase by 1.4 to 4.6. To achieve this, we will employ techniques of visual interactivity and uncertainty visualization.

[The historical daily temperature datasets](historical%20daily%20temperature) were downloaded from [Meteorological Service Singapore website](http://www.weather.gov.sg/home/), consisting of daily mean temperatures recorded for January of the year 1983, 1993, 2003, 2013, and 2023 in weather station Changi.

## 2. Getting Started

### 2.1 Installing and loading the required R packages

In this exercise, we use p_load() of pacman package to load required R packages. The packages that will be used are:

-   *tidyverse*

-   *ggstatsplot*

-   *ggiraph*

```{r}
#| code-fold: true
#| result: false
pacman::p_load(tidyverse, ggiraph, ggstatsplot)
```

### 2.2 Importing the data

::: panel-tabset
## Import

The downloaded datasets consist of five separate CSV files. The code chunk below imports all the five files into R environment by using [read_csv()](https://readr.tidyverse.org/reference/read_delim.html) function of [readr](https://readr.tidyverse.org/) package.

```{r}
#| eval: false
data_1983 <- read_csv("data/DAILYDATA_S24_198301.csv",locale=locale(encoding="latin1"))
data_1993 <- read_csv("data/DAILYDATA_S24_199301.csv",locale=locale(encoding="latin1"))
data_2003 <- read_csv("data/DAILYDATA_S24_200301.csv",locale=locale(encoding="latin1"))
data_2013 <- read_csv("data/DAILYDATA_S24_201301.csv",locale=locale(encoding="latin1"))
data_2023 <- read_csv("data/DAILYDATA_S24_202301.csv",locale=locale(encoding="latin1"))
```

Next, we will merge the data and save the resulting object to an RDS file, which will then be loaded into the working environment.

```{r}
#| eval: false
dataset <- bind_rows(data_1983, data_1993, data_2003, data_2013, data_2023)
write_rds(dataset,"data/dataset.rds")
```

```{r}
dataset <- read_rds("data/dataset.rds")
```

## Preview

```{r}
#| code-fold: true
head(dataset)
```
:::

## 3. Data Wrangling

### 3.1 Variables Selection

We will select our variables of interest from 13 variables and narrow them down to 6 variables. They are: *Year, Month, Day, Mean Temperature (°C), Maximum Temperature (°C), Minimum Temperature (°C)*. Subsequently, we will simplify the variable names for convenience.

::: panel-tabset
## Code

```{r}
# select variables
subset <- dataset %>%
  select('Year','Month','Day','Mean Temperature (°C)','Maximum Temperature (°C)',
         'Minimum Temperature (°C)')

# rename
subset <- subset %>%
  rename('Mean_temp'='Mean Temperature (°C)',
         'Max_temp'='Maximum Temperature (°C)',
         'Min_temp'='Minimum Temperature (°C)')
```

## Preview

```{r}
#| code-fold: true
head(subset)
```
:::
