---
title: "In-Class Exercise 1: Now You See It!"
date: "January 13, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
---

## Loading R packages

In this hands-on exercise, two R packages will be used. They are:

-   [tidyverse](https://www.tidyverse.org/), and

-   [haven](https://haven.tidyverse.org/)

The code chunk used is as follows:

```{r}
pacman::p_load(tidyverse,haven)
```

## Importing PISA data

The code chunk below uses [read_sas()](https://haven.tidyverse.org/reference/read_sas.html) of [haven](https://haven.tidyverse.org/) to import PISA data into R environment.

```{r}
# eval:false
stu_qqq <- read_sas("data/cy08msp_stu_qqq.sas7bdat")
```

The code chunk below uses [filter()](https://dplyr.tidyverse.org/reference/filter.html) of [dplyr](https://dplyr.tidyverse.org/) to filter where country (CNT) is Singapore (SGP).

```{r}
stu_qqq_SG <- stu_qqq %>%
filter(CNT == "SGP")
```

[write_rds()](https://readr.tidyverse.org/reference/read_rds.html) is a function to save an R object in a file using the RDS (R Data Serialization) file format. [read_rds()](https://readr.tidyverse.org/reference/read_rds.html) is a function used to read an R object from a file in the RDS format. RDS is a binary format that allows you to save and load R objects efficiently.

```{r}
write_rds(stu_qqq_SG,"data/stu_qqq_SG.rds")
```

```{r}
stu_qqq_SG <- read_rds("data/stu_qqq_SG.rds")
```
